# Reproducible Research: Peer Assessment 1

Here we will analize data from an unspecified *personal activity monitoring device*. The device collects data a 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Loading and preprocessing the data

```{r}
setwd("~/Desktop/data_science/05_reproduce/assign/RepData_PeerAssessment1/")
# extract and load activity.csv file from activity.zip
act_data <- read.csv(unz("activity.zip", "activity.csv")) 
```

## What is mean total number of steps taken per day?

```{r results='asis'}
act_data.no_na <- act_data[!is.na(act_data$steps),] # remove NAs
steps_per_day = tapply(act_data.no_na$steps, act_data.no_na$date, sum)
steps_per_day[is.na(steps_per_day)] = 0

# create a data frame
mean_total_steps<- round(mean(steps_per_day))
```

The mean total number of steps per day, rounded to the nearest step, is `r mean_total_steps`.

## What is the average daily activity pattern?

```{r}
library(scales)
mean_steps_per_interval = tapply(act_data.no_na$steps, act_data.no_na$interval, mean)
plot(act_data.no_na$interval, act_data.no_na$steps, col = alpha("light blue", 0.25), pch=16)
lines(names(mean_steps_per_interval), mean_steps_per_interval, lwd=3, col="black")

```

## Imputing missing values

Lets assume that missing values from each interval are missing **at random**, since our sample has a large number of points, lets impute the missing values by taking a random sample from other intervals in other days. 

```{r}
set.seed(12345)
missing_set <- act_data[is.na(act_data$steps),]
non_missing_set <- act_data[!is.na(act_data$steps),]

for (i in row.names(missing_set) ) {
    missing_interval = act_data[i,"interval"]
    sample_set <- non_missing_set[non_missing_set$interval==missing_interval,"steps"]
    samp <- sample(sample_set, size=1, replace=TRUE)
    act_data[i, "steps"] = samp
}

# plot the activity levels again to see if there are any differences
mean_steps_per_interval = tapply(act_data$steps, act_data$interval, mean)
plot(act_data$interval, act_data$steps, col = alpha("light blue", 0.25), pch=16)
lines(names(mean_steps_per_interval), mean_steps_per_interval, lwd=3, col="black")


```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
act_data$dayOfWeek = weekdays(as.Date(act_data$date))
act_data$weekend = (act_data$dayOfWeek %in% c("Saturday", "Sunday"))
act_data_weekend = act_data[act_data$weekend == TRUE,]
act_data_weekday = act_data[act_data$weekend == FALSE,]

quartz()
mean_steps_per_interval = tapply(act_data_weekend$steps, act_data_weekend$interval, mean)
plot(act_data_weekend$interval, act_data_weekend$steps, col = alpha("salmon", 0.25), pch=16)
points(act_data_weekday$interval, act_data_weekday$steps, col = alpha("light blue", 0.25), pch=16)
lines(names(mean_steps_per_interval), mean_steps_per_interval, lwd=3, col="red")
mean_steps_per_interval = tapply(act_data_weekday$steps, act_data_weekday$interval, mean)
lines(names(mean_steps_per_interval), mean_steps_per_interval, lwd=3, col="blue")


```